---
title: "AML_Lab4_Group8"
author: "Hyungyum Kim, Milda Poceviciute, Boxi Zhang, Fanny Karelius"
date: "17 october 2018"
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(987654321)
library(readr)
library(kernlab)
library(AtmRay)
```

# Question 2.1

##1
```{r, echo=FALSE}
conf_band <- function(mean, var){
  upp_band <- mean+qnorm(0.975)*sqrt(diag(var))
  low_band <- mean-qnorm(0.975)*sqrt(diag(var))
  return(list(upper=upp_band, lower=low_band))
}
```

For the first exercise, the following Gaussian process regression model will be implemented:

$$ y = f(x) + \epsilon, \ \ \epsilon \sim N(0, \sigma_n^2), \ \  f \sim GP(0, k(x,x^\prime))$$
The function is called posteriorGP and has following inputs:

- X: Vector of training inputs.
- y: Vector of training targets/outputs.
- XStar: Vector of inputs where the posterior distribution is evaluated. i.e. $X_{\star}$.
- hyperParam: Vector with two elements, $\sigma_f$ and $\ell$.
- sigmaNoise: Noise  standard deviation $\sigma_n$.


```{r,eval=TRUE}
SE_kernel <- function(sigmaf, ell){
  val <- function(x1, x2){
    n1 <- length(x1)
    n2 <- length(x2)
    K <- matrix(NA,n1,n2)
    for (i in 1:n2){
      K[,i] <- sigmaf^2*exp(-0.5*( (x1-x2[i])/ell)^2 )
    }
    return(K)
  }
  class(val) <- "kernel"
  return(val)
}

posteriorGP <- function(x, y, xStar, hyperParam, sigmaNoise){
  #### Input:
  #x = train data
  #y = targets
  #xStar = test data
  #hyperParam = list(sigmaF, l)
  #sigmaNoise
  #K covariance for indata, call se_kernel
  y <- as.matrix(y)
  sigmaF <- hyperParam$sigmaF
  l <- hyperParam$l
  kernel <- SE_kernel(sigmaF, l)
  k <- kernel(x, x)
  L <- t(chol(k+sigmaNoise^2*diag(ncol(k))))
  kStar <- kernel(x, xStar)
  alpha <- solve(t(L))%*%(solve(L)%*%y)
  fStar <- t(kStar)%*%alpha #predictive mean
  v <- solve(L)%*%kStar
  V <- kernel(xStar, xStar)-t(v)%*%v #predictive variance
  #log_marg <- -0.5*t(y)%*%alpha-sum(L) - 0.5*length(x)*log(2*pi)
  #### Output:
  #posterior mean, posterior variance
  return(list("pred_mean"=fStar, "pred_var"=V))
}

```


##2

In this part we assume that prior hyper parameters are $\sigma_f = 1$, $\ell = 0.3$, and we update the posterior with observation $(x,y)=(0.4,0.719)$. Here $\sigma_n$ is assumed to be 0.1. The plot below shows the observation, the posterior mean, and the 95% probability bands (point wise) plotted over a grid $[-1,1]$. The probability bands are quite wide for most of the $x$ grid points, except from the location of the observed point $(x,y)$. This observation decreases the uncertainty and the bands become narrower.

```{r}
noise <- 0.1
hyperParam <- list("sigmaF"=1, "l"=0.3)
xGrid <- seq(-1,1,length=50)

obs1 <- list(x=0.4, y=0.719)
post1<-posteriorGP(obs1$x, obs1$y, xGrid, hyperParam, noise)
confband1 <- conf_band(post1$pred_mean, post1$pred_var)

```

```{r, echo=FALSE}
plot(xGrid, post1$pred_mean, type="l", ylim=c(-2,4), ylab="Posterior mean", xlab = "x", main="One observation")
lines(xGrid,confband1$upper, col="blue")
lines(xGrid,confband1$lower, col="blue")
abline(v=0.4,col="red")
legend("topleft",c("posterior mean", "probability bands", "observation"),
       col = c("black","blue", "red"), lty = c(1, 1, 1), lwd=c(1, 1, 1))

```


##3

The posterior is updated with one more observation $(x,y)=(-0.6,-0.044)$. The plot below shows again the observations, the posterior mean, and the 95% (point wise) probability bands over the grid $[-1,1]$. This time the uncertainty is smaller for the two observed $x$ values: the probability bands are narrower there.

```{r}
obs2 <- list(x=c(obs1$x,-0.6), y=c(obs1$y,-0.044))
post2 <- posteriorGP(obs2$x, obs2$y, xGrid, hyperParam, noise)
confband2 <- conf_band(post2$pred_mean, post2$pred_var)

```

```{r, echo=FALSE}
plot(xGrid, post2$pred_mean, type="l", ylim=c(-2,4), ylab="Posterior mean", xlab = "x", main="Two observations")
lines(xGrid,confband2$upper, col="blue")
lines(xGrid,confband2$lower, col="blue")
abline(v=0.4,col="red")
abline(v=-0.6,col="red")
legend("topright",c("posterior mean", "probability bands", "observations"),
       col = c("black","blue","red"), lty = c(1, 1, 1), lwd=c(1, 1, 1))

```


##4

The plot shows the posterior mean and the 95% probability bands after 5 observations. The uncertainty has decreased a lot compared to when we only had one and two observations (as indicated by the decreased width of the probability bands).

```{r}
obs3 <- list(x=c(obs2$x, -1, -0.2, 0.8), y=c(obs2$y, 0.768, -0.94, -0.664))
post3 <- posteriorGP(obs3$x, obs3$y, xGrid, hyperParam, noise)
confband3 <- conf_band(post3$pred_mean, post3$pred_var)

```

```{r, echo=FALSE}
plot(xGrid, post3$pred_mean, type="l", ylim=c(-2,2), ylab="Posterior mean", xlab = "x", main="Five observations")
lines(xGrid,confband3$upper, col="blue")
lines(xGrid,confband3$lower, col="blue")
legend("topright",c("posterior mean", "probability bands"),
       col = c("black","blue"), lty = c(1, 1), lwd=c(1, 1))

```


##5

The plot below shows the posterior mean and the 95% probability bands after 5 observations as in part 4, but changing the hyper parameters to $\sigma_f = 1$, $\ell = 1$. The higher the $\ell$ value, the smoother the posterior mean becomes. In the kernel, the distance between the points is divided by $\ell$. This means that a higher value of this hyper parameter makes the  kernel values converge to $\sigma_f$: the distance between the observations becomes less and less important. Hence, the posterior mean becomes more smooth.

```{r}
hyperParam2 <- list(sigmaF=1, l=1)
post4 <- posteriorGP(obs3$x, obs3$y, xGrid, hyperParam2, noise)
confband4 <- conf_band(post4$pred_mean, post4$pred_var)

```

```{r, echo=FALSE}
plot(xGrid, post4$pred_mean, type="l", ylim=c(-1.5,1), ylab="Posterior mean", xlab = "x", main="sigmaF=1, l=1")
lines(xGrid,confband4$upper, col="blue")
lines(xGrid,confband4$lower, col="blue")
legend("topright",c("posterior mean", "probability bands"),
       col = c("black","blue"), lty = c(1, 1), lwd=c(1, 1))

```


# Question 2.2

For this exercise, data of daily mean temperatures in Tullinge are used.

##1


```{r}
TempTullinge <- read.csv("TempTullinge.csv", header=TRUE, sep=";")

time <- 1:length(TempTullinge$date)
day <- 1:365
fTime <- seq(from=time[1], to=time[length(time)], by = 5)
fDay <- seq(from=day[1], to=day[length(day)], by = 5)
temp <- TempTullinge$temp[fTime]


kernel_func <- SE_kernel(sigmaf=1, ell=0.3)
point1 <- c(1,2)
kernel_point1 <- kernel_func(point1[1], point1[2])
x_vec <- c(1,3,4)
xStar_vec <- c(2,3,4)
k_cov1<-kernelMatrix(kernel = kernel_func, x = x_vec, y = xStar_vec)
```

For the point $x=1, x'=2$, the squared exponential kernel value (with $\sigma_f=1, \ell = 0.3$) was `r kernel_point1`.

For the vectors $X=(1,3,4)^T$ and $X_*=(2,3,4)^T$, the covariance matrix (with $\sigma_f=1, \ell = 0.3$) was 
```{r, echo=FALSE}
k_cov1

```


##2

The following model was used to obtain the posterior mean:
$$temp = f(time) + \epsilon$$
with $\epsilon \sim N(0,\sigma_n^2)$, $f \sim GP(0,k(time,time^{'}))$

$\sigma_n^2$ was obtained by taking the variance of the residuals from a quadratic regression of $temp$ as a function of $time$.

The hyper parameters used in the squared exponential kernel were $\sigma_f=20, \ell=0.2$. A Gaussian process model was fitted on every fifth observation. The posterior mean was calculated by predicting on the training data.

```{r}
lm_fit <- lm(temp~time+time^2, data = TempTullinge)
sigma2n <- var(lm_fit$residuals)
kernel_func2 <- SE_kernel(sigmaf=20, ell=0.2)
gp_model <- gausspr(fTime,temp,data=TempTullinge, kernel=kernel_func2, var=sigma2n)
post_mean <- predict(gp_model, fTime)

```


The plot below shows the observations plotted with a fitted posterior mean. The model fits the data quite well as it predicts the underlying trends (periodicity).

```{r, echo=FALSE}
plot(fTime, temp, pch=20, main="Posterior mean of f", xlab="Time")
lines(fTime,post_mean, col="blue", lwd=2)
legend("bottomright",c("observations", "posterior mean (time)"),
       col = c("black","blue"), pch=c(20,-1), lty = c(0, 1), lwd=c(0, 1))

```

##3

```{r}
x <- scale(fTime)
xs <- scale(fTime)
n <- length(fTime)
Kss <- kernelMatrix(kernel = kernel_func2, x = xs, y = xs)
Kxx <- kernelMatrix(kernel = kernel_func2, x = x, y = x)
Kxs <- kernelMatrix(kernel = kernel_func2, x = x, y = xs)
Covf = Kss-t(Kxs)%*%solve(Kxx + sigma2n*diag(n), Kxs) # Covariance matrix of fStar

```

From the plot below, we see that the probability and prediction bands fit the data quite well. Though, the prediction bands are quite wide.

```{r, echo=FALSE}
# Probability intervals for fStar
plot(fTime, temp, ylim=c(-40,35), pch=20, main="Posterior mean of f", xlab="Time")
lines(fTime,post_mean, col="blue", lwd=2)
lines(fTime, post_mean - 1.96*sqrt(diag(Covf)), col = "red", lwd=2)
lines(fTime, post_mean + 1.96*sqrt(diag(Covf)), col = "red", lwd=2)
# Prediction intervals for yStar
lines(fTime, post_mean - 1.96*sqrt((diag(Covf) + sigma2n)), col = "red", lwd=2, lty=2)
lines(fTime, post_mean + 1.96*sqrt((diag(Covf) + sigma2n)), col = "red", lwd=2, lty=2)
legend("bottomright",c("posterior mean (time)", "probability bands","prediction bands"),
       col = c("blue","red", "red"), lty = c(1, 1, 5), lwd=c(1, 1, 1))

```


##4

The following model was used to obtain the posterior mean:
$$temp = f(day) + \epsilon, \ \ \epsilon \sim N(0,\sigma_n^2), \ \ f \sim GP(0, k(day, day^\prime))$$

$\sigma_n^2$ was obtained by taking the variance of the residuals from a quadratic regression of $temp$ as a function of $day$.

The hyper parameters used in the squared exponential kernel were $\sigma_f=20, \ell=0.2$. A Gaussian process model was fitted on every fifth observation. The posterior mean was calculated by predicting on the training data.

```{r}
fDay_rep <- rep(fDay,6)
lm_fit2 <- lm(temp~fDay_rep+fDay_rep^2)
sigma2n_day <- var(lm_fit2$residuals)
kernel_func3 <- SE_kernel(sigmaf=20, ell=1.2)
gp_model2 <- gausspr(fDay_rep,temp,data=TempTullinge, kernel=kernel_func3, var=sigma2n_day)
post_mean2 <- predict(gp_model2, fDay_rep)

```

```{r, echo=FALSE}
plot(fTime, temp, ylim=c(-25,25), pch=20, main="Posterior mean of f", xlab="Time")
lines(fTime,post_mean, col="blue", lwd=2)
lines(fTime,post_mean2, col="red", lwd=2)
legend("bottomright",c("posterior mean (time)","posterior mean (day)"),
       col = c("blue","red"), lty = c(1,1), lwd=c(1,1))

```

The model with prediction variable $time$ shows a more smooth posterior mean curve than the model with prediction variable $day$.
The posterior mean for $day$ captures the seasonal trend of the temperature (as the $day=1,2,...,365$ variable is replicated 6 times), but the posterior mean for $time$ captures the overall trend of temperature. 
Briefly, it is better to use the $time$ model to see changes overtime and better to use the $day$ model to see general seasonal trends within a year.

##5

The GP with a generalization of the periodic kernel is used in this part to model the temperature against time.

The periodic kernel is given by:
$$k(x,x')=\sigma_f^2\exp{\{-\frac{2\sin^2(pi|x-x'|)/d}{\ell_1^2}\}}\exp{\{-\frac{|x-x'|^2}{2\ell_2^2}\}}$$

The hyper parameters used in the kernel were $\sigma_f=20, \ell_1=1, \ell_2=10, d=365/\text{sd}(time)$.

```{r}
periodic_kernel <- function(sigmaf, ell1, ell2, d){
  val <- function(x1, x2){
    n1 <- length(x1)
    n2 <- length(x2)
    K <- matrix(NA,n1,n2)
    for (i in 1:n2){
      K[,i] <- sigmaf^2*exp(-2*sin(pi*abs(x1-x2[i])/d)^2/ell1^2)*exp(-0.5*abs(x1-x2[i])^2/ell2^2)
    }
    return(K)
  }
  class(val) <- "kernel"
  return(val)  
}

periodic_k <- periodic_kernel(20, 1, 10, 365/sd(time))
gp_model3 <- gausspr(fTime,temp,data=TempTullinge, kernel=periodic_k, var=sigma2n_day)
post_mean3 <- predict(gp_model3, fTime)

```

```{r, echo=FALSE}
plot(fTime, temp, pch=20, main="Posterior mean of f", xlab="Time")
lines(fTime,post_mean, col="blue", lwd=2)
lines(fTime,post_mean2, col="red", lwd=2)
lines(fTime,post_mean3, col="green", lwd=2)
legend("bottomright",c("posterior mean (time)","posterior mean (day)","posterior mean (periodic)"),
       col = c("blue","red","green"), lty = c(1, 1, 1), lwd=c(1, 1, 1))

```

From the plot, it seems that the model with periodic kernel captures more within year seasonality than the model with squared exponential kernel with variable $time$ and also shows more overall trend of temperature than the model with squared exponential kernel with variable $day$. Since the periodic kernel has one more length scale, which controls the correlation between the same day in different years, this result looks reasonable.

# Question 2.3

Banknote fraud data was used to fit a Gaussian process classification model. 1000 observations were used as training data and the rest as test data.

##1

Using the `kernlab` library, a GP classification model was trained on the training data using the covariates $varWave$ and $skewWave$.

```{r, message=FALSE}
fraud_data <- read.csv("https://github.com/STIMALiU/AdvMLCourse/raw/master/GaussianProcess/Code/banknoteFraud.csv", header=FALSE, sep=",")
names(fraud_data) <- c("varWave","skewWave","kurtWave","entropyWave","fraud") 
fraud_data[,5] <- as.factor(fraud_data[,5])

set.seed(111)
SelectTraining <- sample(1:dim(fraud_data)[1], size = 1000, replace = FALSE)
fraud_train <- fraud_data[SelectTraining,]
fraud_test <- fraud_data[-SelectTraining,]
gp_fraud1 <- gausspr(fraud~varWave+skewWave, data=fraud_train)
x1 <- seq(min(fraud_data$varWave),max(fraud_data$varWave),length=100)
x2 <- seq(min(fraud_data$skewWave),max(fraud_data$skewWave),length=100)
fraud_grid <- meshgrid(x1,x2)
grid_points <- data.frame(cbind(c(fraud_grid$x), c(fraud_grid$y)))
names(grid_points) <- c("varWave","skewWave")
probPreds1 <- predict(gp_fraud1, grid_points, type="probabilities")

```

The contours of prediction probability were plotted over a grid of $varWave$ and $skewWave$ values. The training data was overlayed.

```{r, echo=FALSE}
contour(x1,x2,matrix(probPreds1[,1], 100, byrow = TRUE), xlab = "varWave", ylab = "skewWave", main = 'Prob(fraud) - fraud is blue')
points(fraud_train[fraud_train$fraud==0,"varWave"],fraud_train[fraud_train$fraud==0,"skewWave"],col="red",pch=20)
points(fraud_train[fraud_train$fraud==1,"varWave"],fraud_train[fraud_train$fraud==1,"skewWave"],col="blue",pch=20)

conf_mat1<-table("Prediction"=predict(gp_fraud1, fraud_train[,1:2]), "True"=fraud_train$fraud)
accuracy1<-sum(diag(conf_mat1))/sum(conf_mat1)

```
\newpage
The confusion matrix and accuracy rate are given by

```{r, echo=FALSE}
conf_mat1
accuracy1

```


##2

Using the model obtained, predictions were made on the test data. The confusion matrix and accuracy rate was

```{r}
conf_mat2<-table("Prediction"=predict(gp_fraud1, fraud_test[,1:2]), "True"=fraud_test$fraud)
accuracy2<-sum(diag(conf_mat2))/sum(conf_mat2)
conf_mat2
accuracy2
```

The accuracy is `r accuracy2`, which is pretty close to the accuracy rate obtained on the training data. This is a good indication that the classification works well and that the model is not over fitting the training data.

##3

Using all four covariates, a GP classification model was trained on the training data and then predictions were made using the test data. 

```{r, results="hide"}
gp_fraud2 <- gausspr(fraud~., data=fraud_train)
conf_mat3<-table("Prediction"=predict(gp_fraud2, fraud_test), "True"=fraud_test$fraud)
accuracy3<-sum(diag(conf_mat3))/sum(conf_mat3)

```

The confusion matrix and accuracy rate was
```{r, echo=FALSE}
conf_mat3
accuracy3
```

Here, the accuracy is very close to 1, and there was only one misclassification. The result with only two covariates was not bad since the accuracy was `r accuracy2`, but obviously the result is better when all covariates are considered. 

#Appendix

```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```
